Practical consistency management for geographically distributed MMOG servers

### 2011-04-14 trunk/textos/optpaxos/draft.tex

@@

Each player sends its commands to one of the servers\footnote{We consider that each server is replicated and that each player may send his command to one of the replicas, instead of the main server, if that provides a lower delay between the issuing of a command and its in-order delivery.}. A command $C = \{ c_1, c_2, ... \}$ is composed of one or more subcommands, one subcommand per object it affects. We refer to the set of objects affected by command $C$ as $obj(C)$.

@@

To ensure reliability despite server crashes, each server is replicated using state-machine replication, implemented with Paxos. Each player may send its commands to any of the replicas of a server ()
%
Each command is assigned a timestamp and executed against objects in timestamp order. We implement this by assigning each object a logical clock.
If the timestamp of a command $C$ is smaller than the logical clock of the object it affects, $C$ is not executed against the object; otherwise, $C$ is executed and its timestamp is used to set the logical clock of the objects if affects. Therefore, the challenge is how to assign timestamps to commands such that consistency is not violated and commands are not discarded due to stale timestamp values. There are three cases to consider:

@@

The server that received $C$ then forwards it to the group coordinator. In the command, it is informed which objects are affected by it. With this information, then, the coordinator picks the highest value ($TS$) among the logical clocks of the involved objects and assigns it as a timestamp for $C$. The logical clocks of all the affected objects are, then, set to a value higher than $TS$. To do so, an instance of Paxos is run, in order to inform the group about the new command, its timestamp and about the new values for the logical clocks of the affected objects.

@@

We rely on a variation of Skeen's algorithm to implement this case.
The server that receives $C$ forwards $C$ to all the coordinators of the objects affected by $C$. Each one of these coordinators, then, proposes a timestamp for $C$. ``To propose" here means to achieve consensus (with Paxos, for example) in the group regarding $C$'s timestamp proposed by that group. The value proposed by a group coordinator when it initiates the consensus instance in its group corresponds to the highest value among the logical clocks of the objects affected by $C$ AND which are managed by its group.

After a group reaches consensus about its timestamp proposal for $C$, it sends it to the other groups. When a group possesses all the timestamp proposals, it picks the highest among them ($TS$), and assigns it as a timestamp to $C$. For each group whose proposal was not the highest one, the logical clocks of its object(s) affected by $C$ must be increased to a value higher than $TS$, which is done by means of consensus in the group.

@@

The basic idea of the optimistic ordering is the following: assuming that the nodes have a synchronized clock\footnote{We don't require here perfectly synchronized clocks, as the optimistic protocol tolerates mistakes by its very definition. We only need clocks which are synchronized enought, so that our delivery order prediction succeeds and matches the conservative delivery order.}, whenever a server (either coordinator or replica) receives a command $C$ from a client, it immediately applies a provisory (optimistic) timestamp $pTS$ for it, which consists of the current time $now$ plus the estimated communication delay $Dc$ to the coordinator. Therefore, $pTS~=~now~+~Dc$. A wait window of length $W$ is considered, where $W$ is defined as the highest estimated communication delay between the server and each of the other servers of its group. We explain $Dc$ later.

After applying the provisory timestamp to the $C$, the server immediately forwards it to the other servers of that group, and puts $C$ on a list, where it stays until $now > C.pTS + W$, which means that $C$ has been in that list for a time longer than the defined wait window. In the meantime, other commands, sent from other clients and forwarded by other servers, may have been received and also inserted in that list, always sorting by the provisory timestamp of these commands. If $W$ has been correctly estimated as the highest communication delay between the server and each of the other servers, and no message was lost, than all the commands that were supposed to be delivered before $C$ have necessarily been received already. If the same occurred for all the servers, then all of them have received all the commands, and can deliver them in the same order of $pTS$.

However, even if the optimistic order is the same for all the servers, it won't be valid if the conservative order is different from it. That is why we add the value $Dc$ to the provisory timestamp of each command. The conservative ordering is executed by the coordinator as the commands are received by it -- in the order in which they were received. When two servers $S_1$ and $S_2$ forward the commands $C_1$ and $C_2$ to the coordinator, the order in which it will receive them will depend on the communication delay between the coordinator and the sender. Assuming that the clocks are sufficiently synchronized and that the communication delays have been correctly estimated, the coordinator will receive\footnote{It's very unlikely that provisory timestamps won't clash, but it is reasonable to assume that, for commands affecting the same objects, and given a timestamp resolution of a few miliseconds, they will probably have sufficiently different timestamps, so that the receival order at the coordinator matches the one predicted by its replicas.} $C_1$ before $C_2$ iff $C_1.pTS~<~C_2.pTS$. Therefore (and if these optimistic hypotheses prove to be true), the conservative delivery order (based on the order in which the coordinator received the commands) will be the same as the optimistic delivery order (based on the provisory timestamp assigned by the replicas).

Nevertheless, even with a very good delay estimation (e.g. on an environment with a low jitter), there is absolutely no guarantee that the optimistic delivery order will match the conservative one. When it doesn't, it is considered a mistake. Every mistake of the optimistic algorithm -- either a lost command message, or an out-of-order delivery, will cause a rollback of the optimistic state of the objects and re-execution of some of the optimistically delivered commands.

To perform that, we consider that each object has an optimistic delivery queue, $Q_{opt}$. Whenever a command is optimistically delivered, the optimistic state is updated and the command is pushed in the back of $Q_{opt}$. Whenever a command $C_c$ is conservatively delivered, it updates the conservative state of the object and the algorithm checks whether it is the first command in $Q_{opt}$. If it is, $C_c$ is simply removed from $Q_{opt}$ and the execution continues. If it isn't, it means that a command was either lost or was optimistically delivered out of order. We then check whether $Q_{opt}$ contains $C_c$. If it does, it is removed from that list -- if $Q_{opt}$ doesn't contain $C_c$, it was probably lost, or might be the very unlikely case where the conservative delivery happened before the optimistic one, so $C_c$ is stored in a list of possibly delayed optimistic delivery and, if it is ever optimistically delivered, the algorithm will know that it should only discard that command. Then the optimistic state is overwritten with the conservative one and, from that state, all the remaining comands in $Q_{opt}$ are re-executed, leading to a new optimistic state for that object.

>>>>

\newcommand{\code}[1]{\texttt{\small{\textbf{#1}}}}

@@

Each player sends its commands to one of the servers\footnote{We consider that each server is replicated and that each player may send his command to one of the replicas, instead of the main server, if that provides a lower delay between the issuing of a command and its delivery.} -- in the case of avatar based virtual environments, and as an avatar is usually also an object, the server to which a player is connected is his avatar's coordinator. A command $C = \{ c_1, c_2, ... \}$ is composed of one or more subcommands, one subcommand per object it affects. We refer to the set of objects affected by command $C$ as $obj(C)$. Also, we refer to the objects coordinated by a server $S$ as $obj(S)$. Finally, we define $obj(C,S)$ = \{$o$ : $o \in obj(C)$ and $o \in obj(S)$\}.

@@

To ensure reliability despite server crashes, each server is replicated using state-machine replication, implemented with Paxos. Each player may send its commands to any of his server's replicas -- which then forwards it to the server. Each command is assigned a timestamp and executed against objects in timestamp order. We implement this by assigning each object a logical clock. If the timestamp of a command $C$ is smaller than the logical clock of the object it affects, $C$ is not executed against the object; otherwise, $C$ is executed and its timestamp is used to set the logical clock of the objects if affects. Therefore, the challenge is how to assign timestamps to commands such that consistency is not violated and commands are not discarded due to stale timestamp values. There are three cases to consider:

@@

The server that received $C$ then forwards it to the group coordinator. In the command, it is informed which objects are affected by it. With this information, then, the coordinator locks all the involved objects, picks the highest value ($TS$) among their logical clocks and assigns it as a timestamp for $C$. After that, an instance of Paxos is run in order to store the command and its timestamp in the memory of replicas. Upon reaching consensus, the logical clocks of all the affected objects are, then, set to a value higher than $TS$ and they are all unlocked.

@@

\code{(we have to check whether this really works\ldots)}

We rely on a variation of Skeen's algorithm to implement this case. The server that receives $C$ forwards $C$ to all the coordinators of the objects affected by $C$. Each coordinator $S$ then locks all objects in $obj(C,S)$ and proposes a timestamp for $C$. ``To propose" here means to achieve consensus (with Paxos, for example) in the group regarding $C$'s timestamp proposed by that group. The value proposed by a group coordinator when it initiates the consensus instance in its group corresponds to the highest value among the logical clocks of the objects in $obj(C,S)$.

After a group reaches consensus about its timestamp proposal for $C$, it sends it to the other groups. When a group with a coordinator $S$ possesses all the timestamp proposals, it picks the highest among them ($TS$), assigns it as a timestamp to $C$, and unlocks all objects in $obj(C,S)$. For each group whose proposal was not the highest one, the logical clocks of its object(s) affected by $C$ must be increased to a value higher than $TS$, which is done by means of consensus in the group.

@@

The basic idea of the optimistic ordering is the following: assuming that the nodes have a synchronized clock\footnote{We don't require here perfectly synchronized clocks, as the optimistic protocol tolerates mistakes by its very definition. We only need clocks which are synchronized enought, so that our delivery order prediction succeeds and matches the conservative delivery order.}, whenever a server (either coordinator or replica) receives a command $C$ from a client, it immediately applies a provisory (optimistic) timestamp $pTS$ for it, which consists simply of the current value of the server's wallclock, $now$. Therefore, $pTS=now$. A wait window of length $w(S)$ is considered, where $w(S)$ is defined as the highest estimated communication delay between the server $S$ and any of the other servers of its group.

After applying the provisory timestamp to $C$, the server immediately forwards it to the other servers of that group, and puts $C$ on a list, where it stays until $now>C.pTS+w(S)$, which means that $C$ has been in that list for a time longer than the defined wait window. In the meantime, other commands, sent from other clients and forwarded by other servers, may have been received and also inserted in that list, always sorting by the provisory timestamp of these commands. If $w(S)$ has been correctly estimated as the highest communication delay between the server and each of the other servers, and no message was lost, than all the commands that were supposed to be delivered before $C$ have necessarily been received already. If the same occurred for all the servers, then all of them have received all the commands, and can deliver them in the same order of $pTS$.

However, even if the optimistic order is the same for all the servers, it won't be valid if the conservative order is different from it. For that reason, before submitting a command to be conservatively ordered, the coordinator $S_c$ also waits for $w(S_c)$, which is the maximum delay between $S_c$ and any of its replicas. Assuming that the clocks are sufficiently synchronized and that the communication delays have been correctly estimated, then after the time $w(S_c)$ has elapsed, the coordinator will have received all the commands forwarded by any of its replicas (until the instant $t:t=now-w(S_c)$), and it will be able to conservatively order them in the same order of the optimistic delivery, assuming that the final order is the same as the order in which the messages were submitted by the coordinator. Therefore, given the optimistic assumptions we have made, the conservative delivery order (based on the order in which the coordinator FIFO-broadcast the commands to the group) will be the same as the optimistic delivery order (based on the provisory timestamp assigned by the replicas).

\subsection{Recovering from mistakes}

Unfortunately, even with a very good delay estimation (e.g. on an environment with a low jitter), there is absolutely no guarantee that the optimistic delivery order will match the conservative one. When it doesn't, it is considered a \emph{mistake}. Every mistake of the optimistic algorithm -- either a lost command message, or an out-of-order delivery -- will cause a rollback of the optimistic state of the objects and re-execution of some of the optimistically delivered commands.

To perform that, we consider that each object has an optimistic delivery queue, $Q_{opt}$. Whenever a command is optimistically delivered, the optimistic state is updated and the command is pushed in the back of $Q_{opt}$. Whenever a command $C_c$ is conservatively delivered, it updates the conservative state of each object in $obj(C_c)$ and, for each one of them, the algorithm checks whether it is the first command in $Q_{opt}$. If it is, $C_c$ is simply removed from $Q_{opt}$ and the execution continues. If it isn't, it means that $C_c$ was either optimistically delivered out of order, or it was simply never optmistically delivered. It then checks whether $Q_{opt}$ contains $C_c$. If it does, it means the command was optimistically deliverd out of order, and it is removed from the list -- if $Q_{opt}$ doesn't contain $C_c$, it was probably lost\footnote{Also, when $C_c$ is delivered, but it is not in $Q_{opt}$, the remaining possibility is the very unlikely case where the conservative delivery happened before the optimistic one. To handle this case, $C_c$ is stored in a list of possibly delayed optimistic delivery and, if it is ever optimistically delivered, the algorithm will know that it should only discard that command, instead of updating the optimistic state.}. Then, the optimistic state is overwritten with the conservative one and, from that state, all the remaining comands in $Q_{opt}$ are re-executed, leading to a new optimistic state for that object.
